---
title             : "RANKING APPLICATIONS FOR NURSERY SCHOOLS CLASSIFICATION PROBLEM"
shorttitle        : "Ranking applications for nursery schools"

author: 
  - name          : "Viviane Adohouannon"
    affiliation   : "1"
  - name          : "Kate Alexander"
    affiliation   : "1"
  - name          : "Juan Arangote"
    affiliation   : "1"
  - name          : "Dian Azbel"
    affiliation   : "1"
  - name          : "Igor Baranov"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "York University School of Continious Studies"

author_note:
  Group TFG: CSDA1010SUMA18, Lab assignment 1 - Classification Problem

abstract:
   Nursery Database was derived from a hierarchical decision model originally developed to rank applications for nursery schools
  

bibliography      : ["r-references.bib"]

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
linkcolor         : "blue"
class             : "man"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
```

```{r analysis_preferences}
# Seed for random number generation
set.seed(42)
```


# Introduction
## Background
## Objective and Hypothesis
## Assessment of Situation
## Plan

# Data understanding
## Preparation
## Cleaning
## Splitting the dataset into train and test

# Feature engineering

# Modeling
## Decision Tree - Iteration 1
### Decision Tree model fit
### Decision Tree model evaluation

## Random Forest - Iteration 2
### Random Forest model fit
### Random Forest model evaluation

## Random Forest - Iteration 3
### Random Forest model fit
### Random Forest model evaluation

# Dicsussion

# Conclusion

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
